\ttitle

### 2014-01-20 Chapters/Chapter1.tex
Changed objective/subjective to quantitative/qualitative

Besides the recorded data sets we have used artificial data sets to obtain an objective performance measure of the method.
These artificial data sets are modeled to the sets used in \cite{camci2010change,takeuchi2006unifying}.
In \Cref{Chapter5} we discuss the artificial data sets and results in detail.
The real-world recordings and results are discussed in \Cref{Chapter6}.

@@

\section{Thesis structure}\label{sec:intro_thesis_structure}
The structure of this thesis is as follows.
In \Cref{Chapter2} a literature review is provided.
We will look at the different interpretations and implementations for change, novelty, and outlier detection.
Previous work in the field of \gls{har} is discussed and we end with existing applications of \glspl{svm} to change detection.
\Cref{Chapter3} further analyses \glspl{svm} and two implementations of \glspl{oc-svm}.
It relates the properties of the constructed \gls{oc-svm} models to change detection in time series data.
That relation is applied to our problem formulation in \Cref{Chapter4}, where we construct our change detection method \gls{ocs-hats}~\cite{vlasveld2014hats}, based on the work of Camci~\cite{camci2010change}.
In \Cref{Chapter5} we apply the method to artificial data and show the objective performance.
In \Cref{Chapter6} the method is applied to our real-world \gls{almende-data} \cite{vlasveld2014acras}.
In that chapter we show the ability of detecting change in \gls{har} data, recorded by inertial sensors.
This thesis is concluded in \Cref{Chapter7}, in which we reflect on the performed research and state possibilities for feature research.

>>>>

Besides the recorded data sets we have used artificial data sets to obtain an objective, quantitative, performance measure of the method.
These artificial data sets are modeled to the sets used in \cite{camci2010change,takeuchi2006unifying}.
In \Cref{Chapter5} we discuss the artificial data sets and results in detail.
The real-world recordings and results are discussed in \Cref{Chapter6}.

@@

\section{Thesis structure}\label{sec:intro_thesis_structure}
The structure of this thesis is as follows.
In \Cref{Chapter2} a literature review is provided.
We will look at the different interpretations and implementations for change, novelty, and outlier detection.
Previous work in the field of \gls{har} is discussed and we end with existing applications of \glspl{svm} to change detection.
\Cref{Chapter3} further analyses \glspl{svm} and two implementations of \glspl{oc-svm}.
It relates the properties of the constructed \gls{oc-svm} models to change detection in time series data.
That relation is applied to our problem formulation in \Cref{Chapter4}, where we construct our change detection method \gls{ocs-hats}~\cite{vlasveld2014hats}, based on the work of Camci~\cite{camci2010change}.
In \Cref{Chapter5} we apply the method to artificial data and show the quantitative performance.
In \Cref{Chapter6} the method is applied to our real-world \gls{almende-data} \cite{vlasveld2014acras}.
In that chapter we show, by qualitative analysis, the ability of detecting change in \gls{har} data, recorded by inertial sensors.
This thesis is concluded in \Cref{Chapter7}, in which we reflect on the performed research and state possibilities for feature research.

### 2014-01-20 Chapters/Chapter4/data_gathering.tex
Changed objective/subjective to quantitative/qualitative

% ------
\subsection{Artificial data}\label{subsec:data_gathering_artificial}
In order to provide an objective comparison to other methods, we will use artificial data sets which are also used in the earlier work on which \gls{ocs-hats} is based.
These are the data sets used by Takeuchi and Yamanishi \cite{takeuchi2006unifying} and Camci \cite{camci2010change}.
Both construct a collection of one-dimensional time series data according to a second order \gls{ar} model:
\begin{equation}
  x_t = a_1 x_{t-1} + a_2 x_{t-2} + \epsilon_t.
\end{equation}
Over the different data series the mean and variance of the Gaussian random variable $\epsilon_t$ differs and changes at pre-determined change points.
Using this data set an objective quality measure over the change detection methods can be obtained and compared.
All the used data sets are listed and analyzed in \Cref{Chapter5}.

% ------
\subsection{Real-world data}\label{subsec:data_gathering_real_world}
In the second type of data sets we apply our method for change detection and temporal segmentation to real-world data sets.
For our setup we record the activities of humans performed both in- and outdoor in an unknown environment.
Activities performed include sitting, standing, walking, running in a straight and curved line, and walking up- and downstairs.
\gls{ocs-hats} uses the signals from the accelerometer, magnetic field, and rotation sensors.
These time series data are used to detect change points.
A video recording from the performed activity is used to annotate the time series with real change points.
The discovered change points are compared with these annotated change points to give a subjective quality measure.
In \Cref{Chapter6} we give a detailed analysis of the performed activities and the recorded data sets.

>>>>

% ------
\subsection{Artificial data}\label{subsec:data_gathering_artificial}
In order to provide an objective, quantitative, comparison to other methods, we will use artificial data sets which are also used in the earlier work on which \gls{ocs-hats} is based.
These are the data sets used by Takeuchi and Yamanishi \cite{takeuchi2006unifying} and Camci \cite{camci2010change}.
Both construct a collection of one-dimensional time series data according to a second order \gls{ar} model:
\begin{equation}
  x_t = a_1 x_{t-1} + a_2 x_{t-2} + \epsilon_t.
\end{equation}
Over the different data series the mean and variance of the Gaussian random variable $\epsilon_t$ differs and changes at pre-determined change points.
Using this data set an quantitative quality measure over the change detection methods can be obtained and compared.
All the used data sets are listed and analyzed in \Cref{Chapter5}.

% ------
\subsection{Real-world data}\label{subsec:data_gathering_real_world}
In the second type of data sets we apply our method for change detection and temporal segmentation to real-world data sets.
For our setup we record the activities of humans performed both in- and outdoor in an unknown environment.
Activities performed include sitting, standing, walking, running in a straight and curved line, and walking up- and downstairs.
\gls{ocs-hats} uses the signals from the accelerometer, magnetic field, and rotation sensors.
These time series data are used to detect change points.
A video recording from the performed activity is used to annotate the time series with real change points.
The discovered change points are compared with these annotated change points to give a qualitative quality measure.
In \Cref{Chapter6} we give a detailed analysis of the performed activities and the recorded data sets.

### 2014-01-20 Chapters/Chapter5/artificial_data_generation.tex
Changed objective/subjective to quantitative/qualitative

% !TEX root = ../../main.tex
\section{Artificial data generation}\label{sec:artificial_data_generation}
In this section we present four data sets, as used by Camci~\cite{camci2010change} and Takeuchi and Yamanishi \cite{takeuchi2006unifying}, to provide for a objective performance comparison.
All the data sets are modeled by a second-order \gls{ar}-model as in \Cref{eq:artificial_data_sets_model}, where $\epsilon_t$ is a Gaussian distribution modeling the noise with mean $0$ and variance $\sigma^2 = 1$, $a_1 = 0.6$ and $a_2 = -0.5$.
\begin{equation}\label{eq:artificial_data_sets_model}
  x_t = a_1 x_{t-1} + a_2 x_{t-2} + \epsilon_t,
\end{equation}
The length of $x_t$ is $10000$ and change points are generated at each $y \times 10000^\text{th}$ data point, with $y = (1, 2, 3, \dots, 9)$.

>>>>

% !TEX root = ../../main.tex
\section{Artificial data generation}\label{sec:artificial_data_generation}
In this section we present four data sets, as used by Camci~\cite{camci2010change} and Takeuchi and Yamanishi \cite{takeuchi2006unifying}, to provide for a quantitative performance comparison.
All the data sets are modeled by a second-order \gls{ar}-model as in \Cref{eq:artificial_data_sets_model}, where $\epsilon_t$ is a Gaussian distribution modeling the noise with mean $0$ and variance $\sigma^2 = 1$, $a_1 = 0.6$ and $a_2 = -0.5$.
\begin{equation}\label{eq:artificial_data_sets_model}
  x_t = a_1 x_{t-1} + a_2 x_{t-2} + \epsilon_t,
\end{equation}
The length of $x_t$ is $10000$ and change points are generated at each $y \times 10000^\text{th}$ data point, with $y = (1, 2, 3, \dots, 9)$.

### 2014-01-20 Chapters/Chapter5/quality_metrics.tex
Chapter 6 run order
Changed objective/subjective to quantitative/qualitative

% !TEX root = ../../main.tex
\section{Quality metrics}\label{sec:artificial_data_quality_metrics}
For an objective measure of quality of the proposed method applied to the four aforementioned data sets, we employ two metrics, inspired by and slightly altered from Takeuchi and Yamanishi~\cite{takeuchi2006unifying}.
The first metric is the average delay $d_{avg}$, for which we first need to define the notion of delay $d$ for each change:
\begin{equation}\label{eq:delay}
  d(y) = |t_y - t_y^*|,
\end{equation}
where $t_y$ is the time of the closest discovered change point and $t_y^*$ is the true change time.
We can then define the $d_{avg}$ over all the discovered change points $Y$ as:
\begin{equation}\label{eq:average_delay}
  d_{avg}(Y) = \frac{\sum_{y=1}^Y d(y)}{Y}.
\end{equation}
Besides the delay, and the standard deviation $std(d_{avg})$ to measure the spread, which should be low to indicate discovered change points near the actual change point, we need to penalize discovered change points which do not relate to a true change point.
For this we use the \acrlong{far} $\gamma$, expressed as
\begin{equation}\label{eq:false_alarm_rate}
  \gamma(Y) = \frac{Y_{\operatorname*{false}}}{Y}.
\end{equation}

For a high quality method the average delay $d_{avg}$ and \gls{far} $\gamma$ should both be low.
In our results we will also list the parameters used, since these need to be user-defined and are part of the (manual) optimization process.
The parameters are:
\begin{itemize}
  \item \textbf{Window length}: The window length that is processed by the \gls{svdd} algorithm. This is the number of data points that are continuously modeled in the shape in a hypersphere.
  \item \textbf{Sigma of \gls{rbf}}: The value of $\sigma$ for the \gls{rbf} kernel. The effect of this parameter has been discussed in \Cref{subsec:svm_model_parameters}.
  \item \textbf{High threshold}: The $th_\text{high}$ threshold value for the \gls{rt} method described in \Cref{subsec:ratio_thresholding}. A higher value indicates a more precise and robust setting.
  \item \textbf{Low threshold}: The $th_\text{low}$ threshold value for the \gls{rt} method described in \Cref{subsec:ratio_thresholding}. A lower value indicates a more precise and robust setting.
  \item \textbf{Closeness}: In the post-processing phase the detected change points which are within a certain \emph{closeness} range will be merged together. This parameter indicates the minimal number of data points that change points must differ in order to be considered separate detections.
\end{itemize}
For all the sets we have used a value of $C = 0.1$, which indicates that $10\%$ of the data points were considered to be outliers.
This is a normal setting for the \gls{svdd} method.

>>>>

% !TEX root = ../../main.tex
\section{Quality metrics}\label{sec:artificial_data_quality_metrics}
For an quantitative measure of quality of the proposed method applied to the four aforementioned data sets, we employ two metrics, inspired by and slightly altered from Takeuchi and Yamanishi~\cite{takeuchi2006unifying}.
The first metric is the average delay $d_{avg}$, for which we first need to define the notion of delay $d$ for each change:
\begin{equation}\label{eq:delay}
  d(y) = |t_y - t_y^*|,
\end{equation}
where $t_y$ is the time of the closest discovered change point and $t_y^*$ is the true change time.
We can then define the $d_{avg}$ over all the discovered change points $Y$ as:
\begin{equation}\label{eq:average_delay}
  d_{avg}(Y) = \frac{\sum_{y=1}^Y d(y)}{Y}.
\end{equation}
Besides the delay, and the standard deviation $std(d_{avg})$ to measure the spread, which should be low to indicate discovered change points near the actual change point, we need to penalize discovered change points which do not relate to a true change point.
For this we use the \acrlong{far} $\gamma$, expressed as
\begin{equation}\label{eq:false_alarm_rate}
  \gamma(Y) = \frac{Y_{\operatorname*{false}}}{Y}.
\end{equation}

For a high quality method the average delay $d_{avg}$ and \gls{far} $\gamma$ should both be low.
In our results we will also list the parameters used, since these need to be user-defined and are part of the (manual) optimization process.
The parameters are:
\begin{itemize}
  \item \textbf{Window length}: The window length that is processed by the \gls{svdd} algorithm.
  This is the number of data points that are continuously modeled in the shape in a hypersphere.
  \item \textbf{Sigma of \gls{rbf}}: The value of $\sigma$ for the \gls{rbf} kernel. The effect of this parameter has been discussed in \Cref{subsec:svm_model_parameters}.
  \item \textbf{High threshold}: The $th_\text{high}$ threshold value for the \gls{rt} method described in \Cref{subsec:ratio_thresholding}.
  A higher value indicates a more precise and robust setting.
  \item \textbf{Low threshold}: The $th_\text{low}$ threshold value for the \gls{rt} method described in \Cref{subsec:ratio_thresholding}.
  A lower value indicates a more precise and robust setting.
  \item \textbf{Proximity}: In the post-processing phase the detected change points which are within a certain range $\delta$ will be merged.
  This parameter indicates the minimal number of data points that change points must differ in order to be considered separate detections.
\end{itemize}
For all the sets we have used a value of $C = 0.1$, which indicates that $10\%$ of the data points were considered to be outliers.
This is a normal setting for the \gls{svdd} method.

### 2014-01-20 Chapters/Chapter6.tex
Updates
Changed objective/subjective to quantitative/qualitative

In this chapter we discuss the real-world data sets used for this research and the performance of our proposed algorithm on these data sets.
In \Cref{sec:data_sets} we discuss available data sets and argue that these are inappropriate for our research.
It is followed by an overview of the setup we have created to record our own continuous data sets in an unknown environment, both for in- and outdoor activities.
The series of activities performed are discussed in \Cref{sec:data_gathering}.
In that section we show typical plots of the inertial sensor data and stills from the video recordings.
That will illustrate the setting in which the activities are performed and the characteristics of continuous data.
Finally, the results of the algorithm are discussed in \Cref{sec:real_world_results}.
We will use the objective measures, introduced in \Cref{sec:artificial_data_quality_metrics}, and also give subjective claims on the performance of the algorithm applied to the real-world data.
Since the data sets used are real-world recordings, we will put emphasis on the latter type of quality measures \ie the subjective claims.

>>>>

In this chapter we discuss the real-world data sets used for this research and the performance of our proposed algorithm on these data sets.
In \Cref{sec:data_sets} we discuss available data sets and argue that these are inappropriate for our research.
It is followed by an overview of the setup we have created to record our own continuous data sets in an unknown environment, both for in- and outdoor activities.
The series of activities performed are discussed in \Cref{sec:data_gathering}.
In that section we show typical plots of the inertial sensor data and stills from the video recordings.
That will illustrate the setting in which the activities are performed and the characteristics of continuous data.
The results of the algorithm are discussed in \Cref{sec:real_world_results}.
Finally, we provide a list of possible improvements in \Cref{sec:possible_improvements}.

The results are measured with both quantitative measures, as introduced in \Cref{sec:artificial_data_quality_metrics}, and qualitative claims on the performance of the algorithm applied to the real-world data.
Since the data sets used are real-world recordings, we will put emphasis on the latter type of quality measures \ie the qualitative claims.

### 2014-01-20 Chapters/Chapter6/data_sets.tex
Updates
Changed objective/subjective to quantitative/qualitative

With these requirements in mind, we have created the following setup.
All activities were recorded by a smartphone worn on body in the right front pants pocket of the subject.
The used phones are the \textsc{HTC Sensation XE} and \textsc{HTC Desire}, running the \textsc{Android} smartphone operating system.
During the activities the inertial sensor data is recorded using a free application, \textsc{Sensor Logger}~\cite{sensorlogger}.
This application records the sensor data to \textsc{CSV} files, which are then transformed to \textsc{MATLAB} compatible files.
The constructed algorithm processes the data in an online manner (using only current and historic data).
The \textsc{MATLAB} implementation uses the \gls{svdd} library \textsc{dd\_tools} by Tax \cite{Ddtools2013}.
During the activities, the subject is recorded with a video camera to enable manual labeling and determination of the change points.
As illustrated in the top row of \Cref{fig:stills_subject_1_and_2}, the manual labeling is an ambiguous task.
In \Cref{fig:stills_subject_2_change_point} a transition between walking and running is shown.
It is impossible to determine the exact location of the change point and thus objective quality measures (such as the delay) are of less importance.

>>>>

With these requirements in mind, we have created the following setup.
All activities were recorded by a smartphone worn on body in the right front pants pocket of the subject.
The used phones are the \textsc{HTC Sensation XE} and \textsc{HTC Desire}, running the \textsc{Android} smartphone operating system.
During the activities the inertial sensor data is recorded using a free application, \textsc{Sensor Logger}.
This application records the sensor data to \textsc{CSV} files, which are then transformed to \textsc{MATLAB} compatible files.
The constructed algorithm processes the data in an online manner (using only current and historic data).
The \textsc{MATLAB} implementation uses the \gls{svdd} library \textsc{dd\_tools} by Tax \cite{Ddtools2013}.
During the activities, the subject is recorded with a video camera to enable manual labeling and determination of the change points.
As illustrated in the top row of \Cref{fig:stills_subject_1_and_2}, the manual labeling is an ambiguous task.
In \Cref{fig:stills_subject_2_change_point} a transition between walking and running is shown.
It is impossible to determine the exact location of the change point and thus quantitative quality measures (such as the delay) are of less importance.

### 2014-01-20 Chapters/Chapter6/results.tex
Changed objective/subjective to quantitative/qualitative

% !TEX root = ../../main.tex
\section{Results}\label{sec:real_world_results}
In this section we will provide the results of our methods \gls{ocs-hats} applied to the real-world data sets and discuss the quality of the solution.
As stated above, we will use the objective quality metrics as used in \Cref{sec:artificial_data_quality_metrics}.
That will be in the form of a tabular overview of the (manually set) best meta-parameters and results, expressed in \gls{far} $\gamma$ and average delay $d_{avg}$ (following \Cref{sec:artificial_data_results}).
Furthermore, we have analyzed the results from an empirical point of view, since the problem of finding change points in human activities is ambiguous.
For the sake of brevity we excluded those observations from this chapter; they can be found in \Cref{AppendixA}.

\subsection{Objective metrics}
The employed two-stage algorithm requires the setting of user defined parameters.
In \Cref{tab:results_final_real_world} for each run the results are listed.
The right-most column summarizes the overall performance by taking the average over all the runs.
We have used a window length of 50 frames, which represents around $1.16$ seconds of data (the frame-rate of the recordings is not fixed).
The other parameter values, \ie the \gls{rbf} $\sigma$, high and low threshold, and proximity parameter $\delta$, are the result of an educated guess and manual optimization.
In general, a high value for the upper threshold and low value for the lower threshold indicates a robust method, since it indicates that change points are significantly different from homogeneous segments.
The proximity $\delta$ should be low, since that indicates there are few false positives.

@@

\subsection{Subjective observations and remarks}\label{subsec:subjective_results}
It can be ambiguous where to place the (manually annotated) change points, in the case of real-world data.
The objective quality measure can give a distorted view on the performance.
To overcome this, we will state observations and remarks found after close (visual) inspection of the annotated and discovered change points.
In \Cref{fig:plots_all_runs_results} all the results are plotted.
The black solid lines indicate our manually determined change points.
The dashed purple lines indicate the change points as discovered by our method.
We will start with general observations applicable to all the runs.
A more detailed discussion for each run can be found in \Cref{AppendixA}.
Where possible, a graphical illustration will support our observations.

>>>>

% !TEX root = ../../main.tex
\section{Results}\label{sec:real_world_results}
In this section we will provide the results of our methods \gls{ocs-hats} applied to the real-world data sets and discuss the quality of the solution.
As stated above, we will use the quantitative quality metrics as used in \Cref{sec:artificial_data_quality_metrics}.
That will be in the form of a tabular overview of the (manually set) best meta-parameters and results, expressed in \gls{far} $\gamma$ and average delay $d_{avg}$ (following \Cref{sec:artificial_data_results}).
Furthermore, we have analyzed the results from an empirical point of view, since the problem of finding change points in human activities is ambiguous.
For the sake of brevity we excluded those observations from this chapter; they can be found in \Cref{AppendixA}.

\subsection{Quantitative metrics}
The employed two-stage algorithm requires the setting of user defined parameters.
In \Cref{tab:results_final_real_world} for each run the results are listed.
The right-most column summarizes the overall performance by taking the average over all the runs.
We have used a window length of 50 frames, which represents around $1.16$ seconds of data (the frame-rate of the recordings is not fixed).
The other parameter values, \ie the \gls{rbf} $\sigma$, high and low threshold, and proximity parameter $\delta$, are the result of an educated guess and manual optimization.
In general, a high value for the upper threshold and low value for the lower threshold indicates a robust method, since it indicates that change points are significantly different from homogeneous segments.
The proximity $\delta$ should be low, since that indicates there are few false positives.

@@

\subsection{Qualitative observations and remarks}\label{subsec:subjective_results}
It can be ambiguous where to place the (manually annotated) change points, in the case of real-world data.
The quantitative quality measure can give a distorted view on the performance.
To overcome this, we will state observations and remarks found after close (visual) inspection of the annotated and discovered change points.
In \Cref{fig:plots_all_runs_results} all the results are plotted.
The black solid lines indicate our manually determined change points.
The dashed purple lines indicate the change points as discovered by our method.
We will start with general observations applicable to all the runs.
A more detailed discussion for each run can be found in \Cref{AppendixA}.
Where possible, a graphical illustration will support our observations.

### 2014-01-20 Chapters/Chapter7.tex
Changed objective/subjective to quantitative/qualitative

In this research we have shown that it is possible to explicitly obtain a temporal segmentation, expressed as change points, of activities performed by humans.
This was performed in a real-world setting, showing the robustness of the method for noisy data.
Since manual labeling (the annotation of change points on the time series data) of human activities is ambiguous we have put emphasis on the subjective inspection of the results.
The proposed method gives fairly good results, although parameter tuning is a subject of further research.

>>>>

In this research we have shown that it is possible to explicitly obtain a temporal segmentation, expressed as change points, of activities performed by humans.
This was performed in a real-world setting, showing the robustness of the method for noisy data.
Since manual labeling (the annotation of change points on the time series data) of human activities is ambiguous we put emphasis on the qualitative inspection of the results.
The proposed method gives fairly good results, although parameter tuning is a subject of further research.
